# Reproducible Research: Peer Assessment 1
=================================================
```{r echo = TRUE}
#static var for the main directory that contain the files
dataDir <- "data\\activity.csv"

substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

#Clean the data
cleanDates <-  function()
{
  data<- read.csv(dataDir, sep = ",", na.strings = "NA")
  #fix interval format
  data$interval = formatC(data$interval, width = 4, format = "d", flag = "0") 
  # formatDate  
  data$date <- paste(data$date,
                     paste(substr(data$interval,0,2),
                           substrRight(data$interval,2),sep=":"),
                     sep=" ")
  #convert to date  
  data$date <- strptime(data$date ,format="%Y-%m-%d %H:%M")
  data[,c("date","steps")]
}

# create a graphics and return the mean / median by day
numberStepsPerDay <- function(dataset)
{  
  dataset$date = as.Date(dataset$date)   
  stepsPerDay <- aggregate(dataset$steps,list(dataset$date),sum )
  names(stepsPerDay) <- c("date","steps")
  hist( stepsPerDay[,2]
        ,xlab='Total number of step / day'
        ,main='total number of steps taken each day'
        ,col='blue')
  # Mean total number of steps taken per day
  meanDay<- mean(stepsPerDay$steps,na.rm=T)
  # Median total number of steps taken per day
  medianDay<-median(stepsPerDay$steps,na.rm=T)
  return(list(meanDay=meanDay,medianDay= medianDay))
}

# Get the average Daily activity by 5 minutes interval
averageDailyActivity <- function(dataset)
{    
  dataset$date <- strptime(dataset$date ,format="%Y-%m-%d %H:%M")  
  averageSteps<- aggregate(dataset$steps,list(strftime(dataset$date, format="%H:%M")),mean )
  names(averageSteps) = c("interval","steps")
  return(averageSteps)
}

# Create a  time series plot of the 5-minute interval (x-axis) 
# and the average number of steps taken, averaged across all days
plotDates<- function(averageStepsData)  {
  with(averageStepsData,{
       intervals <- strptime(interval, format="%H:%M")
       plot(intervals,steps,type='l',
            main="Average daily activity pattern",     
            ylab="average number of steps taken, averaged across all days",
            xlab="5-minute interval")       
       })
}

# fill NA with the mean for that day
fillEmptyValues <- function(data)
{
  tmpdata= na.omit(data)  
  # Mean total number of steps taken per day  
  meanDay <-aggregate(tmpdata$steps,list(as.Date(tmpdata$date)),mean)
  names(meanDay) = c("dateMerge","stepsMean")
  dataNew <- cbind(data,as.Date(data$date))
  names(dataNew) = c(names(data),"dateMerge")
  dataNew<-merge(x=dataNew,y=meanDay,by.x="dateMerge",by.y="dateMerge")  
  for(i in 1:nrow(dataNew))
  {
    if(is.na(dataNew[i,]$steps))
    {
      dataNew[i,]$steps = dataNew[i,]$stepsMean
    }
  }  
  return(dataNew[,c("date","steps")])
}

# Create the level based on the week day of the date
createLevel <- function(dataNew)
{
  dataNew <-cbind(dataNew,weekdays(as.Date(dataNew$date)))
  names(dataNew) <- c(names(dataNew)[1:2],"weekday")
  dataNew$weekday = as.character(dataNew$weekday)
  dataNew[which(dataNew$weekday == "Saturday" | dataNew$weekday == "Sunday" ),]$weekday="weekend"
  dataNew[which(dataNew$weekday != "weekend"  ),]$weekday="weekday"
  dataNew$weekday = as.factor(dataNew$weekday)
  return(dataNew)
}

# function to create two side graphics baed on the level (Weekday / Weekend)
createLevelGraph <- function(dataNew)
{
  dataNew$date <- strptime(dataNew$date ,format="%Y-%m-%d %H:%M")
  avgNewData <- averageDailyActivity(subset(dataNew,weekday=="weekday"))  
  avgNewDataWeekend <- averageDailyActivity(subset(dataNew,weekday=="weekend"))
  par(mfrow=c(2,1))
  par(mar=c(4,3,2,3))
  with(avgNewDataWeekend,{    
    plot(steps,type='l',
         main="", ylab="",xlab="")})
  title("weekend",col.main="brown")
  with(avgNewData,{    
    plot(steps,type='l',
         main="", ylab="",  
         xlab="Interval")})
  title("weekday",col.main="brown",xlab="interval")
  mtext("Number of steps",2,2)
}
```


## Loading and preprocessing the data
```{r}
  data <- cleanDates()
  dataNaOmit <-  na.omit(cleanDates())
```

## What is mean total number of steps taken per day?

```{r fig.width=7, fig.height=6}  
  res<- numberStepsPerDay(dataNaOmit)
```

- the mean total number of steps is `r  res$meanDay`
- the median total number of steps is `r res$medianDay`


## What is the average daily activity pattern?

```{r fig.width=7, fig.height=6}
averageSteps<- averageDailyActivity(dataNaOmit)
  #5-minute interval with max number of step
  intervalWithMaxNumberSteps <- averageSteps[which(max(averageSteps$steps)==averageSteps$steps),]$interval
  plotDates(averageSteps)
```
- the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is `r intervalWithMaxNumberSteps`


## Imputing missing values
- The total number of missing values in the dataset `r  dim(data[is.na(data$steps)==T,])[1]` 

```{r}
dataNew <- fillEmptyValues(data)
meanAndMedian <- numberStepsPerDay(dataNew)
```
- the mean total number of steps is `r  meanAndMedian$meanDay`
- the median total number of steps is `r meanAndMedian$medianDay`

The mean and median don't differ from the estimates from the first part of the assignment.



## Are there differences in activity patterns between weekdays and weekends?
```{r fig.width=7, fig.height=6}
 # Add levels WeekDays
  dataNew <- createLevel(dataNew)
  #create graphic
  createLevelGraph(dataNew) 
```
